Union and intersection contracts are hard, actually

### 2021-06-09 paper.tex
Rework section 4 (WIP)

In this section, we review some shortcomings of existing work on union and
intersection contracts. We start with the work of Keil and
Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
convincing semantics for union and intersection contracts. We follow with the
work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which builds
upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
uniform blame resolution.

@@

In return, the complexity is offloaded to a separate procedure that must decide
when to raise blame. This complexity is salient in the deterministic, sequential
version\footnote{cf Section 5.1 of \cite{KeilThiemannUnionIntersection}}. There,
the evaluation procedure has to deal again with the problem of eliding contracts
that are not \emph{compatible}\footnote{You can check more on this topic on
Section 5.1 of \cite{KeilThiemannUnionIntersection}} (the ones coming from a
different branch that shouldn't be applied in the current context). In practice,
the evaluation procedure maintains a global environment of contracts that is
extended each time a non-trivial contract (union, intersection or higher-order)
is opened. Then, any contract application must check the presence of the
corresponding label in the environment. But the most surprising\todo{(Yann) find
another word} part is that for contracts attached to a function that is applied,
the whole \emph{context} (the continuation of the application) must be checked
for compatibility with respect to each label in the contract
environment!\todo{(Yann) TBC: find better wording}

@@

% Another difficult issue Keil and Thiemann were able to solve, is the
% fact that user defined contracts and intersection do not usually play nice
% together, as shown on Section \ref{sec:flat-and-inter}.
% They provide two different sets of rules for their calculus,
% the first one is non deterministic, and its purpose is to provide
% semantics that are simple to understand.
% With these rules, the interaction between user defined contracts and
% intersections becomes much simpler, since they are able to check
% both sides of an intersection on separate threads.
% This allows them to be sure that contracts from one branch of the intersection
% won't get blamed while executing the other branch, and vice versa.
%
% As an example, consider the programs shown on Figure
% \ref{fig:kt-non-deterministic}, if the Original program
% gets executed, it will spawn two instances of said program,
% one with the first branch of the intersection (\nickel{3 | Num}),
% and the other one with the second branch (\nickel{3 | Positive}),
% and these will get joined once they reach an equal point, in this case
% it will be \nickel{3}.
%
% \begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Original]
% 3 | Num @& Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=First Program]
% 3 | Num
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Second Program]
% 3 | Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Joined Program]
% 3
% \end{lstlisting}
% \caption{Non deterministic intersection contract opening}
% \label{fig:kt-non-deterministic}
% \end{figure}
%
% The second set of rules is deterministic, and it is intended
% to be simpler to implement.
% However, this second set needs to perform operations
% on the evaluation context, to decide whether or not to evaluate
% a given contract.
% \footnote{You can check more on this topic on Section 5.1 of
% \cite{KeilThiemannUnionIntersection}}
% For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
% when using the value \nickel{(fun x => x) | Str -> Str)}
% inside the execution of contract \nickel{C}, and applying it to \nickel{0},
% the contract runtime would check the context of execution and it would
% deduce that contract \nickel{Str -> Str} does not apply in this case,
% skipping it.
%
% We see two problems here, a minor one is the fact that the two
% different semantics defined for the calculus are really far apart
% one from the other, both on the strategies used for contract resolution,
% as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible.
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}

@@

Williams, Morris, and Wadler, take a different approach on user defined contracts,
they are completely forbidden in their system, explicitly to not encounter
the previously mentioned issue.
This restriction goes against what is usually expected from dynamic checks:
validation of data, and not just inclusion on types.

On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and intersection contracts.
However, one of these properties states the following:

>>>>

In this section, we review some shortcomings of existing work on union and
intersection contracts. We start with the work of Keil and
Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
lean\todo{Not the right word} semantics for union and intersection contracts. We
follow with the work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which
builds upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
uniform blame resolution.

@@

In return, the complexity is offloaded to a separate procedure that must decide
when to raise blame. This complexity is salient in the deterministic, sequential
version\footnote{cf Section 5.1 of \cite{KeilThiemannUnionIntersection}}. There,
the evaluation procedure has to deal again with the problem of eliding contracts
that are not \emph{compatible}\footnote{You can check more on this topic on
Section 5.1 of \cite{KeilThiemannUnionIntersection}} (contracts are incompatible
if they are coming from a different branch and should be ignored). In practice,
the evaluation procedure maintains a global environment of contracts that is
extended each time a non-trivial contract (union, intersection or higher-order)
is opened. Then, any contract application must check the presence of the
corresponding label in the environment. What's more, each time a function with a contract attached is applied,
the whole \emph{context} (that is, the continuation of the application) must be
must be checked for compatibility with respect to each label of the contract
environment. Checking for compatibility involves searching the context for
contract applications.

Keil and Thiemann realizes the tour de force of combining union and intersection
contracts with user-defined contracts. The price to pay, though, is that
implementing their solution in a reasonable way is far from obvious. Not only
when starting from the more abstract, non-deterministic formulation, but even
when starting from the algorithmic version of their system.

@@

% Another difficult issue Keil and Thiemann were able to solve, is the
% fact that user defined contracts and intersection do not usually play nice
% together, as shown on Section \ref{sec:flat-and-inter}.
% They provide two different sets of rules for their calculus,
% the first one is non deterministic, and its purpose is to provide
% semantics that are simple to understand.
% With these rules, the interaction between user defined contracts and
% intersections becomes much simpler, since they are able to check
% both sides of an intersection on separate threads.
% This allows them to be sure that contracts from one branch of the intersection
% won't get blamed while executing the other branch, and vice versa.
%
% As an example, consider the programs shown on Figure
% \ref{fig:kt-non-deterministic}, if the Original program
% gets executed, it will spawn two instances of said program,
% one with the first branch of the intersection (\nickel{3 | Num}),
% and the other one with the second branch (\nickel{3 | Positive}),
% and these will get joined once they reach an equal point, in this case
% it will be \nickel{3}.
%
% \begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Original]
% 3 | Num @& Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=First Program]
% 3 | Num
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Second Program]
% 3 | Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Joined Program]
% 3
% \end{lstlisting}
% \caption{Non deterministic intersection contract opening}
% \label{fig:kt-non-deterministic}
% \end{figure}
%
% The second set of rules is deterministic, and it is intended
% to be simpler to implement.
% However, this second set needs to perform operations
% on the evaluation context, to decide whether or not to evaluate
% a given contract.
% \footnote{You can check more on this topic on Section 5.1 of
% \cite{KeilThiemannUnionIntersection}}
% For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
% when using the value \nickel{(fun x => x) | Str -> Str)}
% inside the execution of contract \nickel{C}, and applying it to \nickel{0},
% the contract runtime would check the context of execution and it would
% deduce that contract \nickel{Str -> Str} does not apply in this case,
% skipping it.
%
% We see two problems here, a minor one is the fact that the two
% different semantics defined for the calculus are really far apart
% one from the other, both on the strategies used for contract resolution,
% as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
% A second, major one,
% is the dependency added over the evaluation context, since, again,
% it makes composing this, with other features desired on programming languages,
% difficult, or even impossible.
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}

@@

As mentioned in Section~\ref{sec:issues-sem}, Williams, Morris, and Wadler,
downright exclude user-defined contracts from the picture. This choice gives
them more freedom in the quest of a more uniform operational semantics. While
this can be a sensible choice (in a gradual typing context e.g.), this is a show
stopper for our initial use-case of the Nickel configuration language, the main
motivation being to use contracts as customizable validation functions.

\ref{sec:issues-sem}
On the semantic side, one of the so-called the sound monitoring properties
linking the operational semantics to the denotational semantics doesn't
correspond to its attached description.
On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and intersection contracts.
However, one of these properties states the following:
